---
title: "Week 1 Problem Based Learning and Practical Solutions"
author: |
  | Jeffrey O. Hanson$^1$
  | $^1$School of Biological Sciences, The University of Queensland, Brisbane, QLD, Australia
  | Correspondance should be addressed to jeffrey.hanson@uqconnect.edu.au
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  rmarkdown::pdf_document:
    toc: true
    toc_depth: 5
    keep_tex: no
    includes:
      in_header: preamble-latex.tex
fontsize: 11pt
documentclass: article
bibliography: ref_lib.bib
csl: reference-style.csl
---

## Problem based learning workshop


## \texttt{R} practical session
### General notes on using the teaching manual 
* network information
* always use Windows 7
	+ Map to L drive: `\\\\sci-tl01.soe.uq.edu.au\\Teaching\\BIOL\\BIOL2006`
	+ Program name: `BIOL2006PracManual_2016.exe`
* sign on
	+ must enter valid id
	+ my sign on is `"Jeffrey"`
* GUI
	+ popup links are bright green
	+ arrows to navigate
	+ home icon is home
	+ R icon saves code to desktop and open up with Rstudio
	+ R code is in blue
	+ comments in black
	+ questions also in green
	+ tooltips appear in brown bar along the bottom of program
	+ minimize question popups by clicking on question again
	+ assessment questions turn black after they're done, but note that you might have to mouse over questions for them to turn black due to bug
	+ students must click logout to store results, they cannot just close the program
	+ students must answer all practice questions to get to final assessment questions
	+ each question in the assessment page refers to a specific page with info on how to answer it
* R scripts
	+ right click on the scripts
	+ open with -> Rstudio
* Potential issues
	* the computers don't associate .R scripts with Rstudio: need to open Rstudio first then open the script
	* attaching datasets: **never to do this even though it looks handy**

### P1. Using R as a calculator
```{r}
1 + 2 
log(10)/( 41 + exp(33.8)) 
x <- (48 + 34)/5.5       # store the result as 'x' 
x         # look at the contents of 'x' 
y <- x + 3 
x*y 
```

### P2. Joining stuff together
```{r}
stuff <- c(1,"cat", 4.5, 6, "elephant") 	# create some "stuff" 
class(stuff) 					# what type is it?  
stuff				# have a look ; note that "4.5" is not a number! 
numbers <- c(1,2.8,3,4.5,12.22,13)	# create "numbers" 
class(numbers) 				# what type is it? 
paste(stuff,numbers)      # join things together as character strings 
3:7      # create a consecutive sequence of numbers 
paste(1:4,stuff,numbers,"whatever!")       
```

### P3. Viewing data
```{r}
women # display the data set called 'women' 
dim(women) # display the numbers of rows and columns 
names(women) # display the column headings 
women$weight # display the column called 'weight' within 'women'
women[4,2] # display row 4 column 2 
plot(women$height, women$weight) # graph 'height' versus 'weight' for 'women' 
attach(women) # declare 'women' to be the default data set 
plot(height, weight, type="l") # graph 'height' vs. 'weight', specifying the line type 
plot(height, weight, type="b", col="red", bty="l", cex=2, xlab="Height", ylab="Weight") 
```

### P4. Importing your own data
```{r}
schoolkids <- data.frame(height=rnorm(100), weight=rnorm(100),
	sex=sample(1:2, 100, replace=TRUE))
schoolkids            # look at the entire data set 
names(schoolkids) 	    # list the variables 
attach(schoolkids)	    # Make it the default data set 
summary(schoolkids) 
schoolkids$sex <- factor(sex) 
# Deduce what each of these commands does: 
schoolkids[2,4] 
schoolkids[,2] 
schoolkids[2,] 
schoolkids[,1:3] 
schoolkids[height>70,]  
```

### P5. Graphing data
```{r}
pairs(schoolkids)  # make pair-wise plots with columns in 'schoolkids' 
plot(height, weight) 		# Plot height (x-axis) against weight (y-axis) 
plot(height,weight,xaxt="n",yaxt="n")   # set up a blank graph with axes  
plot(height,weight,type="n",  
 bty="l",  # Remove the surrounding box 
 xlab="Height in inches",  # set a label for the horizontal axis 
 ylab="Weight in pounds",  # set a label for the vertical axis 
 cex.axis=1.5,  # multiply the axis size by 1.5 
 cex.lab=1.5)  # multiply the axis label size by 1.5 
points(height[sex==1], weight[sex==1])  # add points for the 1st sex 
points(height[sex==2],weight[sex==2],col="red")  # add red points for the 2nd sex
legend(52,160,c("Male","Female"),col=c("black","red"),pch=c(1,1) ) # add  a legend 
```

### P6. Simulating dice throws
```{r}
die <- c(1:6)				# define the possible outcomes for a 6 sided die 
die 						# have a look at your 'die' 
sample(die, 1, replace = TRUE)	# take a random 'throw of the of 'die' 
sample(die, 10, replace = TRUE)		# observe 10 simulated dice throws 
table(sample(die, size = 10000, replace = TRUE))	 
# construct a frequency table for 10,000 dice throws 
mean(sample(die, 10, replace = TRUE))	# observe the mean (average) of ten throws 
# Repeat the previous command several times 
# Observe that the sample mean is "scattered around" a "true mean" of 3.5 
# Try this again, but with sample size increased to 100 
# The "scatter around the true mean" decreases! 
```

### P7. Simulating data within a 'for loop'
```{r}
# In order to visualize the "scatter around a true mean", 
# the following code will generate and store 1000 such means 
results <- vector()				# set up space to hold results 
length(results) <- 1000			# size it to hold exactly 1000 means 
for (i in 1:1000) { # set up a 'for loop' that will iterate 1000 times 
	# save the mean at location 'i' of 'results' 
	results[i] <- mean(sample(die, 10, replace = TRUE)) 
}                # repeat 'for loop' at the next value of 'i' 
hist(results, xlim = c(1,6))	# A histogram of the 1000 means 
# Repeat, but with sample size increased to 100 
# The "scatter around the true mean" decreases! 
```
